Changes in the justification of educational inequalities. The role of perceptions of inequality and meritocracy during the COVID pandemic.

### 2025-01-14 05-discussion.qmd
revpaper -  theory, methods & conclusions

Finally, as our first hypothesis supported, we expected that the relationship between meritocratic perceptions and justification would tend to diminish over time. However, longitudinal evidence shows that the perception of meritocracy linked to effort strengthens its relation with the justification of inequality in education, against the temporal change hypothesis ($H_5$). In other words, meritocracy seems to gain force concerning justifying inequality and not the other way around as we expected given the social context in the post-covid era. An alternative interpretation is that in times of high exposure to risk, people tend to stick more to narratives that provide greater security regardless of the rest of the society - such as individual effort -, which might also be accompanied by lower degrees of solidarity to safeguard the well-being of themselves and their surrounding close peers. Again, this certainly needs further studies that include additional specific conceptual and methodological approaches to this phenomenon.

>>>>

Our first hypothesis was supported, as we anticipated that the relationship between perceptions of meritocracy and the justification of educational inequality would tend to diminish over time. We based this on questions that related directly to experiences during the pandemic, such as receiving economic benefits during the lockdown period. Our findings revealed that those who received benefits justified educational inequality less than those who did not. Moreover, we observed that the connection between perceived meritocracy and the justification of educational inequality weakened for those who received benefits. This evidence partially supports our temporal change hypothesis ($H5$). In other words, the influence of meritocracy on justifying inequality appears to lessen, which aligns with our expectations given the social context in the post-COVID era. One possible interpretation is that, during times of increased risk, people may become more skeptical of narratives emphasizing individual effort over collective efforts. This shift in perspective might also foster stronger solidarity to protect individual and societal wellbeing. However, further studies are necessary to explore this phenomenon using more specific conceptual and methodological approaches.
